Manuscript

### 2018-12-21 manuscript.Rmd
work on manuscript updates

```{r setup, include = FALSE}
# library("papaja")
# install.packages("tinytex")
# tinytex::install_tinytex()
# install.packages("citr")
# devtools::install_github("benmarwick/wordcountaddin", type = "source", dependencies = TRUE)
# update wordcount in Addins
```

@@

G2P designed a set of standards for stratifying their database’s cancer variant interpretations, ranging from preclinical data at the low-evidence end (level D) to clinically actionable interpretations at the high-evidence end (level A). To test only these clinically actionable associations, we filtered the G2P dataset for only level-A G2P interpretations, of which there were 1,296 (see Methods). All compounds screened within the GDSC, CTRP, and CCLE datasets are considered “small molecules” and are respectively linked to Compound (CID) numbers used for indexing in the PubChem database (https://pubchem.ncbi.nlm.nih.gov/). As a result, non-small molecules, which include proteins and biologics (i.e. monoclonal antibodies), that had level-A G2P evidence could not be included in our comparisons. Thus, G2P interpretations for which the compound did not have a matching CID code were excluded from further analysis as were interpretations that did not have an associated compound, resulting in removal of 402 interpretations (Table 1). It should be noted that cisplatin, a small molecule, was excluded due to its being indexed with a ChEMBLdb identifier (https://www.ebi.ac.uk/chembl/), which we did not manually re-index to a CID to avoid accidental mischaracterization.

After filtering, 894 level-A G2P interpretations were carried forward in subsequent analyses, constituting 57 unique drugs and 34 unique genes in 88 distinct level-A combinations (Figure 1A). The number of level-A interpretations per drug ranged from 1 to 176 (M = 15.7, SD = 34.6), from 1 to 264 (M = 26.2, SD = 61.6) per gene, and for each level-A drug-gene combination, the range was 1 to 103 (M = 10.1, SD = 19.8). These unique drugs, genes, and combinations are hereafter referred to as G2P drugs, G2P genes, and G2P associations. Of the 57 unique G2P drugs, 37 of them were screened in at least one of the drug screen datasets, 14 were screened in two of the three datasets, and only three drugs were screened in all three datasets. The remaining 20 un-screened G2P drugs suggests an incompleteness of these drug screens, though these projects frequently release new screening data that may include these drugs in future releases.

Using cell-line-specific point mutation annotations from the CCLE, we compared z-score-transformed area under the dose-response curve (AUC) distributions between mutant and wildtype cell lines for all G2P genes for each G2P drug tested in the CCLE, CTRP, and GDSC drug screens (Figures 1B and 2A). At uncorrected $\alpha$ = 0.05, Wilcoxon rank-sum tests revealed that in CCLE, the AUC distributions of mutant and wildtype groups for 10 out of 170 (5.88%) tested drug-gene combinations were significantly different, 91 out of 918 (9.91%) were significantly different for CTRP, and 61 out of 850 (7.18%) for GDSC. We observe little concordance between the number of interpretations for each G2P association and the significance of mutation status on AUC for said G2P association, suggesting that mutation status alone is not a robust predictor of clinically actionable genetic targets. This led us to examine the correlations between AUC and gene expression and AUC and copy number, two other genomic features that are known to associate with cancer and drug response.

@@

We compared the distribution of level-A G2P associations (Figure 1A) and the associated p-values from the mutation status comparisons (Figure 1B) with a null distribution of p-values derived from all pairwise combinations of G2P genes and the G2P drugs tested in each dataset, less any of the 88 distinct combinations with G2P interpretations, i.e. the non-gray squares in the right heatmap of Figure 1A (Figure 3A). To formally test concordance between clinically actionable variants and drug screen results, we conducted an independent, two-group t-test to compare the distributions of p-values of drug-gene combinations that either were or were not represented with level-A evidence in G2P, which indicated that the distributions of p-values did not significantly differ for CCLE [t(9) = 0.25, p = 0.81], CTRP [t(51) = 0.85, p = 0.40], nor GDSC [t(47) = -0.56, p = 0.58]  (Table 2). This further supports that mutation status is not a robust genomic feature for predicting therapeutic responses.

Similar tests were conducted for the distributions of Pearson correlation coefficients (r) between the number of level-A G2P associations and gene expression (RPKM) and copy number (log2 ratio) of the gene in the association (Figures 3B-C). Example r derivations are shown in Figures 2B-C. Absolute values of r were used to simplify extreme values in both directions for testing. The difference in the CCLE |r| distributions for both gene expression [t(8) = 1.43, p = 0.19] and copy number [t(8) = 2.48, p = 0.23] were non-significant, which is likely explained by the small sample size since only five G2P drugs were screened in the CCLE dataset. For both the CTRP and GDSC screens, |r| values for gene expression correlations were significantly greater for the level-A G2P associations than the null distribution [tCTRP(49) = 3.79, pCTRP < 0.001; tGDSC(44) = 2.97, pGDSC = 0.005]. The |r| distributions using copy number were also significantly different for CTRP [t(49) = 2.48, p = 0.017] but not for GDSC [t(44) = 1.95, p = 0.058].

In all three datasets for both level-A G2P associations and null distribution associations, there are more significant |r| values when using gene expression as the genomic feature than when using copy number (Table 2), which corresponds with previous work showing that gene expression is an informative molecular feature when assessing drug sensitivity (Jang et al., 2014; Sirota et al., 2011). Additionally, in all cases, there are more significant r values when using copy number than there are significant p-values from the mutation status comparisons. Though observational, these results suggest an order of predictive robustness for these three genomic features when used in isolation: Gene expression > copy number > mutation status. Additionally, across all three genomic features, CTRP consistently identified a higher percentage of its possible level-A G2P associations than GDSC, and similarly, GDSC identified a higher percentage than CCLE. In the CCLE-GDSC debate, inconsistency between the datasets was attributed to the groups using two very different assays for generating AUC measurements though it was undecided which was more accurate (Haibe-Kains et al., 2013; Jang et al., 2014; Weinstein & Lorenzi, 2013). Our results provide very preliminary evidence that GDSC is more accurate than CCLE in regards to reproducing clinical associations. However, CTRP may be even more accurate than both GDSC and CCLE.

@@

In the 894 level-A G2P interpretations, there was a wide variation in the specificity of cancer description, ranging from “Waldenström macroglobulinemia” and “hyper eosinophilic advanced syndrome” on the more detailed end to “cancer” on the broad end. Similarly, the drug screen datasets had a wide range in lineage specificity, e.g. CCLE, GTRP, and GDSC all had lineages labeled “leukemia” and “T-cell childhood acute lymphocytic leukemia” and the CRISPR dataset had cell lines labeled “Epstein-Barr virus-related Burkitt lymphoma” and “lymphoma.” To more easily conduct comparative analyses in a cancer-specific manner, we developed a more general lineage grouping method derived from the Human Disease Ontology identity codes (DOIDs, http://disease-ontology.org/) (Schriml et al., 2019) assigned to each cell line, yielding 27 unique lineage groupings. This custom lineage grouping was included in our harmonized cancer cell line database (see METHODS).

In the level-A G2P interpretations, there were 102 unique drug-gene-lineage combinations (DGLs), representing seven of our curated cancer lineages: cancer, breast cancer, leukemia, lung cancer, ovarian cancer, skin cancer, and thyroid cancer. CCLE had data for nine DGLs, CTRP had data for 61, and GDSC had data for 53, yielding a total of 73 unique DGLs across all three datasets.

At an alpha of 0.05, no DGLs were identified in the CCLE screen when using mutation status as a correlate (n = 7), but CTRP and GDSC identified six and five, respectively, in skin cancer, lung cancer, and cancer. These significant associations did not hold after Bonferroni correction in CTRP (n = 48, $\alpha$adjusted = 0.001) or GDSC (n = 39, $\alpha$adjusted = 0.001).

When using gene expression as the genomic feature, three CCLE DGLs had significant correlations at an adjusted alpha of 0.006 (n = 9). Two of these three DGLs were also significantly correlated with copy number (n = 9, $\alpha$adjusted = 0.006).

In CTRP, AUC of 19 DGLs significantly correlated with gene expression, 12 of which were significant after correction (n = 61, $\alpha$adjusted = 0.0008). Copy number correlations were significant for 12 DGLs, with three DGLs significant post-correction (n = 61, $\alpha$adjusted = 0.0008).

AUC of 16 DGLs in GDSC significantly correlated with gene expression. At an adjusted alpha of 0.0009 (n = 53) eight of them were significant. Fewer DGLs were significant when correlating AUC with copy number with 6 significant at an alpha of 0.05 and 1 significant at an adjusted alpha of 0.0009 (n = 53).

Interestingly, when we test the effect of mutation status on drug response for the DGLs derived from all pairwise combinations of G2P drugs and genes… (Figure ##1).

@@

The CRISPR screen from the Broad Institute reports gene dependency using CERES scores (Meyers et al., 2017), which are generated from sgRNA depletion scores and eliminate bias arising from the effect of copy number variation on Cas9 DNA cleavage. The lower the CERES score, the higher the likelihood that the gene is essential in the associated cell line. Scores are scaled per cell line such that a score of 0 is the median effect of nonessential genes and -1 is the median effect of common core essential genes. Of the 34 G2P genes, only G6PD was not targeted by an sgRNA in the CRISPR dataset.

When we don’t account for lineage, Wilcoxon rank-sum tests revealed that mutations in eight of the 33 (24.2%) G2P genes significantly correlated with CERES score ($\alpha$ = 0.05): KRAS (p < 0.001), BRAF (p < 0.001), BRCA2  (p < 0.001), CYP19A1 (p = 0.004), TSC2 (p = 0.005), JAK2 (p = 0.031), DPYD (p = 0.042), and UGT1A1 (p = 0.048). Gene expression (RPKM) of 17 (51.5%) G2P genes significantly correlated with CERES score, as did copy number (log2-ratio) for 14 (42.4%) G2P genes (Figure ##2). For the purpose of comparison to the drug screen data, if we treat CRISPR knockdown like a small-molecule drug, then these results are similar to the results from CCLE, CTRP, and GDSC (Table 2).

@@

In a lineage-agnostic context of the CRISPR screen, ERBB2 gene essentiality score had a significant negative correlation with ERBB2 gene expression (p < 10-12, r = -0.31) and ERBB2 copy number (p < 10-15, r = -0.34) (Figure ##2). Similarly, MET gene essentiality also correlated with both MET gene expression (p < 10-12, r = -0.31) and copy number (p < 10-8, r = -0.25). Notably, BRCA1 also significantly correlated with both gene expression (p < 0.001, r = 0.16) and copy number (p < 0.01, r = 0.14). BRCA2 significantly correlated with all three of the genomic features tested: gene expression (p < 0.05, r = 0.10), copy number (p < 10-7, r = 0.23), and mutation status (p < 0.001).

@@

@@

Level-A G2P interpretations correspond to clinical evidence that suggests efficacy of gene targets and/or drugs, and this work questions whether these relationships manifest themselves in in vitro drug screens. We have demonstrated that on its own, mutation status is not a robust predictor of drug efficacy and that copy number and gene expression fare better as predictors. However, for these three genomic features in all three datasets, only 11.1% (mutation status, 1 of 9, CCLE) to 47.9% (gene expression, 23 of 48, CTRP) (Table 2), of the level-A G2P drug-gene associations in the respective dataset were significant at an uncorrected $\alpha$ = 0.05. This suggests that while the drug screens successfully identify some clinically relevant drug-gene associations, many associations are also missed, which raises the question of the extent to which clinical researchers can rely on in vitro drug screens when attempting to develop and select therapeutics for cancer patients.

@@

The CCLE provided mutation annotations in 19,280 genes across 1,596 cell lines (dataset: depmap_18Q4_mutation_calls.csv, accessed 14 November 2018). CCLE called substitutions using MuTect [@cibulskis2013] and annotated variants using Oncotator [@ramos2015a] and indels using Indelocator (https://software.broadinstitute.org/cancer/cga/indelocator). We filtered mutation calls for point mutations, defined as single-nucleotide insertions, deletions, and substitutions, regardless of result, i.e. frameshift or missense mutation. For this analysis, cell lines harboring non-silent point mutations were considered “mutant” for the gene in question. All other mutations and genes without annotation in the Mutation Annotation Format (MAF) file were considered “wildtype.” “Mutant” and “wildtype” calls were binarily encoded per-gene/per-cell line, regardless of quantity of harbored mutations per gene in a given cell line.

@@

The CCLE reports cell line sensitivity to small molecules as an activity area measured above the dose-response curve (ActArea), while both CTRP and GDSC report small molecule activity as area under the dose-response curve (AUC). Since ActArea is reported with a scale of 0 to 8, we converted CCLE’s ActArea values to AUC via:

@@

For Figure 1, all data were imported and manipulated using pandas (v0.23.0) [@mckinney2011] and plotted using seaborn (v0.90) [@waskom2012], which sat on top of matplotlib (v2.2.2) [@hunter2007] and was executed on Python (v2.7.15) (Python Software Foundation, https://www.python.org/) in an IPython (v5.5.0) [@perez2007] kernel within a localhost Jupyter (v5.2.2) notebook [@kluyver2016]. All other analyses were conducted in R (v3.5.0) (R Foundation for Statistical Computing, https://www.R-project.org/) in an RStudio environment (v1.1.447) [@rstudio2012]. All other plots were drawn using the ggplot2 (v3.0.0) R package [@wickham2016].

>>>>

```{r setup, include=FALSE}
library(tidyverse)
library(ggpubr)
library(papaja)
# install.packages("tinytex")
# tinytex::install_tinytex()
# install.packages("citr")
# devtools::install_github("benmarwick/wordcountaddin", type = "source", dependencies = TRUE)
# update wordcount in Addins
```

@@

```{r g2p-data, include=FALSE}
ccl_converter <- read.delim("./data_munging/cell_line_database_v1_20180911.tsv", row.names = 1, sep = "\t", header = TRUE)
lineage_converter <- unique(ccl_converter[, c("grouped_general_doid", "group_general_lineage_name")])
colnames(ccl_converter)[colnames(ccl_converter) == "Broad_ID"] <- "DepMap_ID"

g2p <- read.delim("./data_munging/g2p_full_level_A_dataset_NEW_DEC_2_2018_withGroupedLineages.csv", sep = "\t", header = TRUE)
g2p[, 1:2] <- NULL
g2p <- merge(g2p, lineage_converter, by = "grouped_general_doid", all.x = TRUE)

g2p$Drug <- tolower(g2p$Drug)
g2p[g2p == -1] <- NA
g2p$Ref[g2p$Ref == "None"] <- NA
g2p$Alt[g2p$Alt == "None"] <- NA
g2p$Drug_Gene <- ifelse(is.na(g2p$Drug) | is.na(g2p$Gene), NA, paste0(g2p$Drug, "_", g2p$Gene))
g2p$Drug_Gene_Lin <- ifelse(is.na(g2p$Drug) | is.na(g2p$Gene) | is.na(g2p$group_general_lineage_name), NA, paste0(g2p$Drug, "_", g2p$Gene, "_", g2p$group_general_lineage_name))
g2p$position <- paste0("g.chr", g2p$Chromosome, ":", g2p$End, g2p$Ref, ">", g2p$Alt)
g2p$position <- ifelse(grepl("NA", g2p$position), NA, g2p$position)
g2p_filt <- subset(g2p, grepl("^CID", g2p$DrugID))
g2p_dgls <-  unique(g2p_filt$Drug_Gene_Lin[!is.na(g2p_filt$Drug_Gene_Lin)])
g2p_druggenes <- unique(g2p_filt$Drug_Gene[!is.na(g2p_filt$Drug_Gene)])
g2p_drugs <- unique(sapply(strsplit(g2p_dgls, "_", fixed = TRUE), function(x) x[1]))
g2p_genes <- unique(sapply(strsplit(g2p_dgls, "_", fixed = TRUE), function(x) x[2]))
g2p_lins <- unique(sapply(strsplit(g2p_dgls, "_", fixed = TRUE), function(x) x[3]))

# Order and genes taken from pink heatmapo from Daniel's analyses
g2p_genes <- c("EGFR", "KRAS", "ERBB2", "ERBB3", "MET", "ALK", "ROS1", "PDGFRA", "ABL1", "KIT", "PDGFRB", "PDGFB", "BRAF", "G6PD", "RET", "BRCA1", "BRCA2", "UGT1A1", "TSC1", "TSC2", "AKT1", "IDH2", "DPYD", "ESR1", "CYP19A1", "TPMT", "FLT3", "NPM1", "DNMT3A", "PML", "PTCH1", "JAK2", "MYD88", "MGMT")

g2p_nonCID <- subset(g2p, !grepl("^CID", g2p$DrugID))

# print("Drug-Gene")
g2p_druggene_tally <- g2p_filt %>% group_by(Drug_Gene) %>% tally() %>% drop_na()
nrow(g2p_druggene_tally)
range(g2p_druggene_tally$n)
mean(g2p_druggene_tally$n)
sd(g2p_druggene_tally$n)

# print("Drug")
g2p_drug_tally <- g2p_filt %>% group_by(Drug) %>% tally() %>% drop_na()
range(g2p_drug_tally$n)
mean(g2p_drug_tally$n)
sd(g2p_drug_tally$n)

# print("Gene")
g2p_gene_tally <- g2p_filt %>% group_by(Gene) %>% tally() %>% drop_na()
range(g2p_gene_tally$n)
mean(g2p_gene_tally$n)
sd(g2p_gene_tally$n)

# print("Drug-Gene-Lineage")
g2p_dgl_tally <- g2p_filt %>% group_by(Drug, Gene, Drug_Gene, group_general_lineage_name) %>% tally() %>% drop_na()
nrow(g2p_dgl_tally)
length(unique(g2p_dgl_tally$Drug))
length(unique(g2p_dgl_tally$Gene))
unique(g2p_filt$Phenotype.Description)
```

```{r table-g2p-nonCID, echo=FALSE, results='asis'}
summ <- g2p_nonCID %>% group_by(Drug, DrugID) %>% tally() %>% summarize("Compound identifier" = DrugID, "Number of G2P interpretations" = n, "Proportion of excluded interpretations (%)" = printnum(n / 402 * 100))
colnames(summ)[colnames(summ) == "Drug"] <- "Compound"
summ$Compound[summ$Compound == "egfr"] <- "EGFR"
summ$Compound[summ$Compound == "mab"] <- "mAb"

apa_table(summ, caption = "Distribution of level-A G2P drugs excluded from analyses",
          note = "mAb = monoclonal antibody; NA indicates G2P interpretation does not have an associated compound")
```

G2P designed a set of standards for stratifying their database’s cancer variant interpretations, ranging from preclinical data at the low-evidence end (level D) to clinically actionable interpretations at the high-evidence end (level A). To test only these clinically actionable associations, we filtered the G2P dataset for only level-A G2P interpretations, of which there were `r formatC(nrow(g2p), big.mark = ",", big.interval = 3L)` (see Methods). All compounds screened within the GDSC, CTRP, and CCLE datasets are considered “small molecules” and are respectively linked to Compound (CID) numbers used for indexing in the PubChem database (https://pubchem.ncbi.nlm.nih.gov/). As a result, non-small molecules, which include proteins and biologics (i.e. monoclonal antibodies), that had level-A G2P evidence could not be included in our comparisons. Thus, G2P interpretations for which the compound did not have a matching CID code were excluded from further analysis as were interpretations that did not have an associated compound, resulting in removal of `r nrow(g2p_nonCID)` interpretations (Table\ \@ref(tab:table-g2p-nonCID)). It should be noted that cisplatin, a small molecule, was excluded due to its being indexed with a ChEMBLdb identifier (https://www.ebi.ac.uk/chembl/), which we did not manually re-index to a CID to avoid accidental mischaracterization.

After filtering, `r nrow(g2p_filt)` level-A G2P interpretations were carried forward in subsequent analyses, constituting `r length(g2p_drugs)` unique drugs and `r length(g2p_genes)` unique genes in `r length(g2p_druggenes)` distinct level-A combinations (*Figure 1A*). The number of level-A interpretations per drug ranged from `r range(g2p_drug_tally$n)[1]` to `r range(g2p_drug_tally$n)[2]` ($M =$ `r printnum(mean(g2p_drug_tally$n))`, $SD =$ `r printnum(sd(g2p_drug_tally$n))`), `r range(g2p_gene_tally$n)[1]` to `r range(g2p_gene_tally$n)[2]` ($M =$ `r printnum(mean(g2p_gene_tally$n))`, $SD =$ `r printnum(sd(g2p_gene_tally$n))`) per gene, and for each level-A drug-gene combination, the range was `r range(g2p_druggene_tally$n)[1]` to `r range(g2p_druggene_tally$n)[2]` ($M =$ `r printnum(mean(g2p_druggene_tally$n))`, $SD =$ `r printnum(sd(g2p_druggene_tally$n))`). These unique drugs, genes, and combinations are hereafter referred to as G2P drugs, G2P genes, and G2P associations. Of the `r length(g2p_drugs)` unique G2P drugs, 37 of them were screened in at least one of the drug screen datasets, 14 were screened in two of the three datasets, and only three drugs were screened in all three datasets. The remaining 20 un-screened G2P drugs suggests an incompleteness of these drug screens, though these projects frequently release new screening data that may include these drugs in future releases.

```{r}
ccle_signif_g2p_grid <- readRDS("./data_munging/rds/ccle_signif_g2p_grid.rds")
ctrp_signif_g2p_grid <- readRDS("./data_munging/rds/ctrp_signif_g2p_grid.rds")
gdsc_signif_g2p_grid <- readRDS("./data_munging/rds/gdsc_signif_g2p_grid.rds")
```

Using cell-line-specific point mutation annotations from the CCLE, we compared z-score-transformed area under the dose-response curve (AUC) distributions between mutant and wildtype cell lines for all G2P genes for each G2P drug tested in the CCLE, CTRP, and GDSC drug screens (*Figures 1B and 2A*). At uncorrected $\alpha = 0.05$, Wilcoxon rank-sum tests revealed that in CCLE, the AUC distributions of mutant and wildtype groups for `r nrow(filter(ccle_signif_g2p_grid, p < 0.05))` out of `r nrow(ccle_signif_g2p_grid)` (`r printnum(nrow(filter(ccle_signif_g2p_grid, p < 0.05)) / nrow(ccle_signif_g2p_grid) * 100)`%) tested drug-gene combinations were significantly different, `r nrow(filter(ctrp_signif_g2p_grid, p < 0.05))` out of `r nrow(ctrp_signif_g2p_grid)` (`r printnum(nrow(filter(ctrp_signif_g2p_grid, p < 0.05)) / nrow(ctrp_signif_g2p_grid) * 100)`%) were significantly different for CTRP, and `r nrow(filter(gdsc_signif_g2p_grid, p < 0.05))` out of `r nrow(gdsc_signif_g2p_grid)` (`r printnum(nrow(filter(gdsc_signif_g2p_grid, p < 0.05)) / nrow(gdsc_signif_g2p_grid) * 100)`%) for GDSC. We observe little concordance between the number of interpretations for each G2P association and the significance of mutation status on AUC for said G2P association, suggesting that mutation status alone is not a robust predictor of clinically actionable genetic targets. This led us to examine the correlations between AUC and gene expression and AUC and copy number, two other genomic features that are known to associate with cancer and drug response.

@@

We compared the distribution of level-A G2P associations (*Figure 1A*) and the associated p-values from the mutation status comparisons (*Figure 1B*) with a null distribution of p-values derived from all pairwise combinations of G2P genes and the G2P drugs tested in each dataset, less any of the 88 distinct combinations with G2P interpretations, i.e. the non-gray squares in the right heatmap of *Figure 1A (Figure 3A*). To formally test concordance between clinically actionable variants and drug screen results, we conducted an independent, two-group t-test to compare the distributions of p-values of drug-gene combinations that either were or were not represented with level-A evidence in G2P, which indicated that the distributions of p-values did not significantly differ for CCLE [t(9) = 0.25, p = 0.81], CTRP [t(51) = 0.85, p = 0.40], nor GDSC [t(47) = -0.56, p = 0.58]  (Table 2). This further supports that mutation status is not a robust genomic feature for predicting therapeutic responses.

Similar tests were conducted for the distributions of Pearson correlation coefficients (r) between the number of level-A G2P associations and gene expression (RPKM) and copy number (log2 ratio) of the gene in the association (*Figures 3B-C*). Example r derivations are shown in *Figures 2B-C*. Absolute values of r were used to simplify extreme values in both directions for testing. The difference in the CCLE |r| distributions for both gene expression [t(8) = 1.43, p = 0.19] and copy number [t(8) = 2.48, p = 0.23] were non-significant, which is likely explained by the small sample size since only five G2P drugs were screened in the CCLE dataset. For both the CTRP and GDSC screens, |r| values for gene expression correlations were significantly greater for the level-A G2P associations than the null distribution [tCTRP(49) = 3.79, pCTRP < 0.001; tGDSC(44) = 2.97, pGDSC = 0.005]. The |r| distributions using copy number were also significantly different for CTRP [t(49) = 2.48, p = 0.017] but not for GDSC [t(44) = 1.95, p = 0.058].

In all three datasets for both level-A G2P associations and null distribution associations, there are more significant |r| values when using gene expression as the genomic feature than when using copy number (Table 2), which corresponds with previous work showing that gene expression is an informative molecular feature when assessing drug sensitivity [@jang2014; @sirota2011]. Additionally, in all cases, there are more significant r values when using copy number than there are significant p-values from the mutation status comparisons. Though observational, these results suggest an order of predictive robustness for these three genomic features when used in isolation: Gene expression > copy number > mutation status. Additionally, across all three genomic features, CTRP consistently identified a higher percentage of its possible level-A G2P associations than GDSC, and similarly, GDSC identified a higher percentage than CCLE. In the CCLE-GDSC debate, inconsistency between the datasets was attributed to the groups using two very different assays for generating AUC measurements though it was undecided which was more accurate[@haibe-kains2013; @jang2014; @weinstein2013]. Our results provide very preliminary evidence that GDSC is more accurate than CCLE in regards to reproducing clinical associations. However, CTRP may be even more accurate than both GDSC and CCLE.

@@

In the `r nrow(g2p_filt)` level-A G2P interpretations, there was a wide variation in the specificity of cancer description, ranging from “Waldenström macroglobulinemia” and “hyper eosinophilic advanced syndrome” on the more detailed end to “cancer” on the broad end. Similarly, the drug screen datasets had a wide range in lineage specificity, e.g. CCLE, GTRP, and GDSC all had lineages labeled “leukemia” and “T-cell childhood acute lymphocytic leukemia” and the CRISPR dataset had cell lines labeled “Epstein-Barr virus-related Burkitt lymphoma” and “lymphoma.” To more easily conduct comparative analyses in a cancer-specific manner, we developed a more general lineage grouping method derived from the Human Disease Ontology identity codes (DOIDs, http://disease-ontology.org/) [@schriml2019] assigned to each cell line, yielding 27 unique lineage groupings. This custom lineage grouping was included in our harmonized cancer cell line database (see METHODS).

In the level-A G2P interpretations, there were `r nrow(g2p_dgl_tally)` unique drug-gene-lineage combinations (DGLs), representing seven of our curated cancer lineages: cancer, breast cancer, leukemia, lung cancer, ovarian cancer, skin cancer, and thyroid cancer. CCLE had data for nine DGLs, CTRP had data for 61, and GDSC had data for 53, yielding a total of 73 unique DGLs across all three datasets.

At an $\alpha =  0.05$, no DGLs were identified in the CCLE screen when using mutation status as a correlate (n = 7), but CTRP and GDSC identified six and five, respectively, in skin cancer, lung cancer, and cancer. These significant associations did not hold after Bonferroni correction in CTRP ($n = 48$, $\alpha_{adjusted} = 0.001$) or GDSC ($n = 39$, $\alpha_{adjusted} = 0.001$).

When using gene expression as the genomic feature, three CCLE DGLs had significant correlations at an adjusted $\alpha =  0.006$ ($n = 9$). Two of these three DGLs were also significantly correlated with copy number ($n = 9$, $\alpha_{adjusted} = 0.006$).

In CTRP, AUC of 19 DGLs significantly correlated with gene expression, 12 of which were significant after correction ($n = 61$, $\alpha_{adjusted} = 0.0008$). Copy number correlations were significant for 12 DGLs, with three DGLs significant post-correction ($n = 61$, $\alpha_{adjusted} = 0.0008$).

AUC of 16 DGLs in GDSC significantly correlated with gene expression. At an adjusted $\alpha$ of 0.0009 ($n = 53$) eight of them were significant. Fewer DGLs were significant when correlating AUC with copy number with 6 significant at an $\alpha$ of 0.05 and 1 significant at an adjusted $\alpha$ of 0.0009 ($n = 53$).

Interestingly, when we test the effect of mutation status on drug response for the DGLs derived from all pairwise combinations of G2P drugs and genes… (*Figure ##1*).

@@

The CRISPR screen from the Broad Institute reports gene dependency using CERES scores [@meyers2017], which are generated from sgRNA depletion scores and eliminate bias arising from the effect of copy number variation on Cas9 DNA cleavage. The lower the CERES score, the higher the likelihood that the gene is essential in the associated cell line. Scores are scaled per cell line such that a score of 0 is the median effect of nonessential genes and -1 is the median effect of common core essential genes. Of the 34 G2P genes, only G6PD was not targeted by an sgRNA in the CRISPR dataset.

When we don’t account for lineage, Wilcoxon rank-sum tests revealed that mutations in eight of the 33 (24.2%) G2P genes significantly correlated with CERES score ($\alpha = 0.05$): KRAS ($p < 0.001$), BRAF ($p < 0.001$), BRCA2  ($p < 0.001$), CYP19A1 ($p = 0.004$), TSC2 ($p = 0.005$), JAK2 ($p = 0.031$), DPYD ($p = 0.042$), and UGT1A1 ($p = 0.048$). Gene expression (RPKM) of 17 (51.5%) G2P genes significantly correlated with CERES score, as did copy number (log2-ratio) for 14 (42.4%) G2P genes (Figure ##2). For the purpose of comparison to the drug screen data, if we treat CRISPR knockdown like a small-molecule drug, then these results are similar to the results from CCLE, CTRP, and GDSC (Table 2).

@@

In a lineage-agnostic context of the CRISPR screen, ERBB2 gene essentiality score had a significant negative correlation with ERBB2 gene expression ($p < 10^{-12}$, $r = -0.31$) and ERBB2 copy number ($p < 10^{-15}$, $r = -0.34$) (*Figure ##2*). Similarly, MET gene essentiality also correlated with both MET gene expression ($p < 10^{-12}$, $r = -0.31$) and copy number ($p < 10^{-8}$, $r = -0.25$). Notably, BRCA1 also significantly correlated with both gene expression ($p < 0.001$, $r = 0.16$) and copy number ($p < 0.01$, $r = 0.14$). BRCA2 significantly correlated with all three of the genomic features tested: gene expression ($p < 0.05$, $r = 0.10$), copy number ($p < 10^{-7}$, $r = 0.23$), and mutation status ($p < 0.001$).

@@

**Should we do this?**

@@

Level-A G2P interpretations correspond to clinical evidence that suggests efficacy of gene targets and/or drugs, and this work questions whether these relationships manifest themselves in in vitro drug screens. We have demonstrated that on its own, mutation status is not a robust predictor of drug efficacy and that copy number and gene expression fare better as predictors. However, for these three genomic features in all three datasets, only 11.1% (mutation status, 1 of 9, CCLE) to 47.9% (gene expression, 23 of 48, CTRP) (Table 2), of the level-A G2P drug-gene associations in the respective dataset were significant at an uncorrected $\alpha = 0.05$. This suggests that while the drug screens successfully identify some clinically relevant drug-gene associations, many associations are also missed, which raises the question of the extent to which clinical researchers can rely on in vitro drug screens when attempting to develop and select therapeutics for cancer patients.

@@

To preclude variation in genomic feature calls across the datasets, we used identical annotations from CCLE for all cell lines screened in the three datasets. The CCLE provided mutation annotations in 19,280 genes across 1,596 cell lines (dataset: depmap_18Q4_mutation_calls.csv, accessed 14 November 2018). CCLE called substitutions using MuTect [@cibulskis2013] and annotated variants using Oncotator [@ramos2015a] and indels using Indelocator (https://software.broadinstitute.org/cancer/cga/indelocator). We filtered mutation calls for point mutations, defined as single-nucleotide insertions, deletions, and substitutions, regardless of result, i.e. frameshift or missense mutation. For this analysis, cell lines harboring non-silent point mutations were considered “mutant” for the gene in question. All other mutations and genes without annotation in the Mutation Annotation Format (MAF) file were considered “wildtype.” “Mutant” and “wildtype” calls were binarily encoded per-gene/per-cell line, regardless of quantity of harbored mutations per gene in a given cell line.

@@

The CCLE reports cell line sensitivity to small molecules as an activity area measured above the dose-response curve (ActArea), while both CTRP and GDSC report small molecule activity as area under the dose-response curve (AUC). Since ActArea is reported with a scale of 0 to 8, we converted CCLE’s ActArea values to AUC via

@@

For *Figure 1*, all data were imported and manipulated using pandas (v0.23.0) [@mckinney2011] and plotted using seaborn (v0.90) [@waskom2012], which sat on top of matplotlib (v2.2.2) [@hunter2007] and was executed on Python (v2.7.15) (Python Software Foundation, https://www.python.org/) in an IPython (v5.5.0) [@perez2007] kernel within a localhost Jupyter (v5.2.2) notebook [@kluyver2016]. All other analyses were conducted in R (v3.5.0) (R Foundation for Statistical Computing, https://www.R-project.org/) in an RStudio environment (v1.1.447) [@rstudio2012]. All other plots were drawn using the ggplot2 (v3.0.0) R package [@wickham2016].
